Application of a Cut-Cell Immersed Boundary Method for Wildfire Simulation over Complex Terrain

### 2020-10-30 Vanella_Article_Atmosphere_2020.tex
Update on Overleaf.

% Authors, for the paper (add full first names)
\Author{Marcos Vanella $^1$, Kevin McGrattan $^1$, Randall McDermott $^1$, Glenn Forney $^1$, and William Mell $^2$}

% Authors, for metadata in PDF
\AuthorNames{Marcos Vanella, Kevin McGrattan, Randall McDermott, Glenn Forney, and William Mell}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad National Institute of Standards and Technology, Gaithersburg, Maryland, USA \\
$^{2}$ \quad U.S. Forest Service, Seattle, Washington, USA}

@@

\abstract{A method for the large-eddy simulation (LES) of wildfire spread over complex terrain is presented. In this scheme, a cut-cell immersed boundary method (CC-IBM) is used to render the complex terrain, defined by a tessellation, on a rectilinear Cartesian grid. Discretization of scalar transport equations for chemical species is done via a finite volume scheme on cut-cells defined by the intersection of the terrain geometry and the Cartesian cells. Momentum transport and heat transfer close to the immersed terrain are handled using dynamic wall models and a direct forcing immersed boundary method. Further, atmospheric wind conditions are specified using a mean forcing concept. Fire spread is modeled using one of three basic approaches: (1) representing the vegetation as a collection of Lagrangian particles, (2) representing the vegetation as a semi-porous boundary, and (3) representing the fire spread using a level set method in which the fire spreads as a function of terrain slope, vegetation type, and wind speed. Several test and validation cases are reported to demonstrate the capabilities of this novel wildfire simulation methodology.}

@@

% Motivation.
In the last decades, wildland and wildland-urban interface (WUI) fires have received substantial attention due to their destructive nature and extensive cost in lives and property~\cite{thomas_2017,mcdermott_2019,boghani_2019,richards_2020}. Over the years, modeling and simulation activities of such phenomena have been tailored to the needs of planners, scientists and first responders, and by the available computing capabilities.
% Wildfire modeling in the literature.
A large body of literature can be found on the subject. See for example reviews~\cite{Papadopoulos_2011,Bakhshaii_2019,mcdermott_2019} on classical and new models with associated software. Due to inherent complex nature of the problem, most models of wildfires have historically relied on simplified field-tested rules and correlations.

Among these, the fire spread rate model by Rothermel~\cite{Rothermel:1972} assumes quasi-steady state surface fire conditions and takes as inputs fuel properties, wind, terrain slope and moisture that can be measured in situ.
Fuel models have been cataloged~\cite{Anderson:1982}, and their input parameters defined in a manner the user can select and combine the most appropriate listed fuels for her case of study. Rothermel's head fire rate of spread formula is a widely used, and has been implemented in several simulation tools~\cite{Finney:FARSITE,Bova:IJWF2015,mcgratta_2013,Coen:2,Coen:2015,LAUTENBERGER_2013,Coen:2013,Mandel:2009,Mandel:2011,Mandel:2014,Kochanski:2016}, enabling a range of planning and operational forecasting capabilities.

% Level set as a simplified fire spread model, operational forecasting.
Evolution of wild land fires is usually tied to a fire front, which can have a complex shape moving/deforming with the local flame spread velocities. Knowing the location of this fire front and local properties of the fire, fuel and terrain allows, by use of simplified models like FARSITE~\cite{Finney:FARSITE}, to predict local ignition and fire spread. When modeling the problem numerically, identifying such fire front at a given moment in time requires a mathematical representation and an associated discretization or mesh. The Lagrangian treatment of the motion of a fire front requires specifying a set of control markers and connectivity that define and move with such front. Therefore, the fire front is represented by this discrete mesh. For example, if the domain of motion is a surface terrain, this Lagrangian mesh can be a polygon that moves with the front along this surface~\cite{Finney:FARSITE,Bova:IJWF2015}.
On the other hand, an Eulerian representation of motion allows for the mesh to be fixed in space (i.e. in a terrain a polygonal patch mesh of said terrain). Then, the location of the fire front can be portrayed implicitly by using a scalar field defined in the fixed two dimensional mesh. This scalar field is called level set function and numerical methods that evolve it in time are called Level Set Methods (LSM)~\cite{Sethian:1999,Osher:2006}. Some wildfire solvers that use LSM to track the fire front are described in references~\cite{coen_2013,Bova:IJWF2015,mcgratta_2013,LAUTENBERGER_2013}. A practical comparison of the two techniques as implemented in WFDS~\cite{Mell:IJWF2007} and FARSITE~\cite{Finney:FARSITE} is provided in reference~\cite{Bova:IJWF2015}.

% Resolving the flow features on terrains using LES over complex terrains. Describe what we do, FDS, cut-cell schemes, immersed boundary methods.
As the rate of spread of the front line is dependent in the local velocity on empirical fire spread models like Rothermels, a better local resolution of the velocity field around complex terrains should be beneficial to the accuracy of the overall model. In reality, besides the obstructive character of the terrain itself, compounded effects due to the fire affect the local velocity field. \textbf{Some of these are ...}.
The de facto physics-based technique for practical simulation of outdoor flows is Large Eddy Simulation (LES)~\cite{coen_2013,mcgratta_2013}. Within FDS~\cite{mcgratta_2013},  the filtered Navier-Stokes equations for buoyant, combusting flows are evolved using a Low Mach approximation. Chemical species are handled using scalar transport equations. Energy transport is accounted for by means of the solution of radiative transport equations and a thermodynamic divergence proxy. \textbf{Comment on WFDS for LOF. The work of Mell et al.~\cite{Mell:IJWF2007}.}.

% Unstructured geometries, capability can be tuned for use with LOF and fire spread models.
% What has been done, cut-cell scheme IBM, connection to LS and Rothermels, wind.
Lately, substantial work has been done to add to FDS the option of simulating flows around unstructured geometries. These are represented by surface triangulations which is a typical representation in computer aided design (CAD) and simulation software. This added flexibility is still compliant with the structured cartesian grids and differentiation schemes within FDS. We implemented a cut-cell scheme~\cite{berger_2016} to advance the scalar transport and energy of multicomponent mixtures on the polyhedra next to the geometries. These polyhedrons arise from geometry intersection with the cartesian fluid grid. As geometry boundaries are not changed, the method has the property of exact volume and area conservation. Velocities next to geometries are reconstructed using an immersed boundary method (IBM)~\cite{fadlun_2000,balaras_2004} coupled with an equilibrium wall model for turbulent flows~\cite{mcdermo_2018}. Variables related to IBM forcing and heat transfer modeling at the wall are sampled through interpolation and a normal probe approach~\cite{balaras_2004}. Also, radiation transport boundary conditions have been adapted to this technique.
For fire simulation over complex terrains a range of options in terms of physics fidelity of the model are being developed, in particular, in the representation of fire spread. For fast simulation of outdoor fires, this new capability has been interfaced with a Level Set scheme~\cite{Bova:IJWF2015} and Rothermels fire spread model. The wind speed input for Rothermels model results from the local velocity over the terrain which is affected by fire buoyancy and flow evolution. These are in turn also influenced by wind atmospheric conditions, accounted for within simulations using a mean wind forcing concept.

% Outline of the manuscript.
In the following section a description on the mathematical model for fire over terrain is provided. The terrain description and model discretization is provided in section~\ref{sec:terraindisc}. An account of the fire spread model is given in section~\ref{sec:firespread}, as well as the wind atmospheric boundary conditions in section~\ref{sec:wind}. Numerical experiments testing the methodology on a simple example and a real case of fire in Cogoleto, Italy are provided in section~\ref{sec:numexp}. Finally a summary of this work is outlined in section~\ref{sec:summary}.

\section{Mathematical model} \label{sec:matmodel}

@@

Consider a mixture composed of $N$ chemical species $\alpha$, moving on a fixed point $\mathbf{x}$ in space with a mass weighted average velocity $\mathbf{u}(\mathbf{x},t)$. If $\rho(\mathbf{x},t)$ is the mixture density, the mass fraction for species $\alpha$ is $Y_\alpha = \rho_\alpha / \rho$, where $\rho_\alpha(\mathbf{x},t)$ is the species mass density and $\rho = \sum \rho_\alpha$. The scalar transport and momentum equations take the form:
\begin{eqnarray}
   \frac{\partial \rho Y_\alpha}{ \partial t} + \nabla \cdot ( \rho Y_\alpha  \mathbf{u} ) &=& - \nabla \cdot \mathbf{J_{d}}_\alpha + \dot{m}_\alpha'''  +
    \dot{m}_{b,\alpha}'''\; , \; \alpha=1,\dots,N \label{eqn:spectran} \\
    \frac{\partial \mathbf{u}}{\partial t} - \mathbf{u} \times \boldsymbol{\omega} + \nabla H - \tilde{p} \, \nabla \left( 1/\rho\right) &=&
    \frac{1}{\rho} \left[ (\rho-\rho_0) \mathbf{g} + \mathbf{f}_{b} + \nabla \cdot \boldsymbol{\tau} \right] \label{eqn:momtran}
\end{eqnarray}
where $\mathbf{J_{d}}_\alpha=- \rho D_\alpha \boldsymbol{\nabla} Y_\alpha$ are the diffusive fluxes for component $\alpha$. Fick's Law for binary diffusion respect to a background species is assumed, $D_\alpha$ is the diffusivity of $\alpha$ respect to the background species.  To enforce realizability, $\sum_\alpha \mathbf{J_{d}}_\alpha = 0$, we lump errors in diffusive transport into the most abundant species locally \cite{McDermott:2015}. The volumetric combustion source term for species $\alpha$ is $\dot{m}_\alpha'''(\mathbf{x},t)$  and $\dot{m}_{b,\alpha}'''$ is the contribution to species $\alpha$ from subgrid particle gasification. In the momentum equation the head is $H=|\mathbf{u}|^2/2 + \tilde{p}/\rho$, $\tilde{p}$ is the perturbation pressure, and $\rho_0(z)$ is an equlibrium densit.  The gravity vector is $\mathbf{g}=(0,0,g_z)$, $\mathbf{f}_{b}$ is the particle force term, and $\boldsymbol{\tau}$ is a tensor accounting for molecular and subgrid turbulent stresses parameterized by an effective eddy viscosity model. This last model is employed by default in large-eddy simulation (LES) of outdoor fires. Details of the model equations and their derivation are found in \cite{mcgratta_2013tr}.

Conservation of energy is achieved by forcing flow field to obey the following thermodynamic divergence constraint:
%
\begin{eqnarray}
    ( \nabla \cdot \mathbf{u} )^{th} &=&
    \left[ \frac{1}{\rho c_p T} - \frac{1}{\bar{p}} \right]
    \frac{\partial \bar{p}}{\partial t} + \frac{w \rho_0 g_z}{\rho c_p T} \nonumber \\
    &+& \frac{1}{\rho c_p T} \left[ \dot{q}''' + \dot{q}_b''' - \nabla \cdot \dot{\mathbf{q}}'' - \mathbf{u} \cdot \nabla (\rho h_s) \right] \nonumber \\
    &+& \frac{1}{\rho} \sum_\alpha \left( \frac{\overline{W}}{W_\alpha} - \frac{h_{s,\alpha}}{c_p T} \right) \left[ \dot{m}_\alpha'''  + \dot{m}_{b,\alpha}''' - \nabla \cdot (\mathbf{J_{d}}_\alpha) - \mathbf{u} \cdot \nabla (\rho Y_\alpha) \right] \label{eq:divth}
\end{eqnarray}
%
where $h_s$ is the local sensible enthalpy, $\dot{q}''', \dot{q}_b'''$ are heat release rates due to combustion and particle gasification, $\dot{\mathbf{q}}''$ is the heat flux, sum of conduction, convection and radiation, and $T$ is the local temperature. The mixture specific heat at constant pressure and molecular weight are $c_p=\sum_{\alpha =1}^N{c_{p,\alpha} Y_\alpha}$ and $\overline{W}=\left(\sum_{\alpha =1}^N{Y_\alpha /W_\alpha} \right)^{-1}$ respectively, built from the individual species counterparts as defined. Finally, $w,\rho_0,g_z$ on the stratification term are respectively, the local vertical velocity, density for standard conditions (depends on vertical coordinate), and vertical gravity acceleration. This thermodynamic divergence is derived from and acts as a proxy for the sensible enthalpy evolution equation~\cite{mcdermo_2014}. Equation (\ref{eq:divth}) is derived by factoring the divergence from the sensible enthalpy equation and applying the ideal gas law.  Given a background pressure and a local mass density the local temperature is computed from the ideal gas law as $T = \bar{p} \overline{W} / (\rho R) $.

% Fire spread model equations.
As commented in the introduction, the location of a fire front can be defined implicitly along the terrain surface through the zero of a two dimensional level set scalar function $\phi(x,y,t)$. This implicit form can be used to define and advection equation for the level set function:
%
\begin{equation}
   \frac{D \phi(x,y,t)}{D t} =  \frac{\partial \phi}{\partial t} + \mathbf{R}_{uv} \cdot \nabla (\phi) = 0 \label{eqn:lset}
\end{equation}
%
where $\mathbf{R}_{uv}=(R_u,R_v)$ is the rate of spread vector along the terrain plane. The magnitude of the rate of spread vector is computed by $|\mathbf{R}_{uv}|=R_0 \left(1 + \sqrt{(\boldsymbol{\phi}_W+\boldsymbol{\phi}_S) \cdot (\boldsymbol{\phi}_W+\boldsymbol{\phi}_S) } \right)$, being $R_0$ the zero velocity and zero slope rate of spread (baseline, dependent of fuel properties), and vectors $\boldsymbol{\phi}_W,\boldsymbol{\phi}_S \ge 0$ are computed using local wind, slope through empirical rules~\cite{Wilson:1980}. See Appendix 1 of reference~\cite{Bova:IJWF2015} for details on these calculations, as implemented in WFDS-LS and in this work.

@@

% How do we describe terrains - geometries. Methodology to get from DEM a geometry.
%
In figure~\ref{Fig:figure_1}a, a terrain is represented by its surface triangulation within an FDS cartesian mesh. Some cartesian cells and faces belonging to this mesh are transversed by the terrain surface. The remainder polyhedra and polygons that lay on the gas side of this intersected cartesian entities are called cut-cells and cut-faces. They discretize the fluid domain in the region surrounding the solid terrain. A detail of this unstructured mesh is. shown in figure~\ref{Fig:figure_1}b. The resulting cut-cells and cut-faces are of varied sizes and shapes, and situations where more than one cut-cell per cartesian cell are not uncommon. As described later, small cells pose a temporal stability constraint in explicit time integration schemes that needs to be addressed.
It is noted that a two level of refinement grid hierarchy emerges. The coarse level is defined by the Cartesian entities, whereas the fine level is defined by the cut-cell or unstructured counterparts. Methods that solve discrete model equations on these grids are called cut-cell or embedded boundary methods~\cite{berger_2016}. Reliably defining the cut-cell mesh geometric properties and topology is a complex problem on itself, omitted here in order to keep the discussion succinct.
%
\begin{figure}[h]
   \centering
   \includegraphics[trim = 12mm 42mm 8mm 25mm, clip,width=1\linewidth]{./figures/SketchFig1.pdf}
   \put(-305,-5){(a)}
   \put(-95,-4){(b)}
   \caption{Sketch of regular Cartesian and cut-cell grids around terrain: (a) An unstructured cut-cell region is defined on the gas phase side around a terrain $\mathbf{T}$. (b) Detail showing a slice representative of cut-cells and regular gas cells, terrain triangulation and two refinement level interpretation.}
   \label{Fig:figure_1}
\end{figure}
%

@@

The equations for chemical species transport are discretized using the finite volume method (FV)~\cite{eymard_2000,leveque_2002}. See figure~\ref{Fig:figure_2}a. Integrating equation~\eqref{eqn:spectran} over a cut-cell $ii$ with control volume $\Omega_{ii}$:
%
\begin{equation}
 \int_{\Omega_{ii}} {\frac{\partial \rho Y_\alpha}{\partial t}} d \Omega + \int_{\Omega_{ii}} { \boldsymbol{\nabla} \cdot  \left(  \rho Y_\alpha \mathbf{u} \right)
      } d \Omega  = -\int_{\Omega_{ii}} { \boldsymbol{\nabla} \cdot \left(  \mathbf{J_{d}}_\alpha  \right)  } d \Omega + \int_{\Omega_{ii}} {( \dot{m}_\alpha'''+\dot{m}_{b,\alpha}''') } d \Omega \label{eq:intconvdiff}
\end{equation}
%
For a control volume that does not change with time, the time derivative and source terms are approximated by
%
\begin{eqnarray}
  \int_{\Omega_{ii}} {\frac{\partial \rho Y_\alpha}{\partial t}} d \Omega & = & \frac{\partial}{\partial t} \int_{\Omega_{ii}} {\rho Y_\alpha} d \Omega
  = \frac{\partial \widetilde{\rho \: Y_\alpha }_{ii}}{\partial t} V_{ii} \\
  \int_{\Omega_{ii}} {( \dot{m}_\alpha''' +\dot{m}_{b,\alpha}''')} d \Omega & = & (\widetilde{ \dot{m}_\alpha''' }_{ii}+ \widetilde{\dot{m}_{b,\alpha}'''}_{ii}) V_{ii} \label{eq:intcons}
\end{eqnarray}
%
where $V_{ii}$ is the volume of cell $ii$ and tildes imply cell averages. In the following the notation is simplified by dropping tildes, observing that quantities are cell or face averaged. Consider the FV discretization of the diffusive term (eqn.~\eqref{eq:intconvdiff}) on cut-cell $ii$ of figure~\ref{Fig:figure_2}a:
%
\begin{equation}
\int_{\Omega_{ii}} { \boldsymbol{\nabla} \cdot \left(  \mathbf{J_{d}}_\alpha  \right)  } d \Omega =
    \int_{\partial \Omega_{ii}} { \left( - \rho D_\alpha \boldsymbol{\nabla} Y_\alpha \right) \cdot \hat{\mathbf{n}}_{ii} } \: d \partial \Omega = \sum^{nf_c}_{k=1}
    \left( - \rho D_\alpha \boldsymbol{\nabla} Y_\alpha \right)_k \cdot \hat{\mathbf{n}}_{ii,k} \: A_k \label{eq:discfvdiffcc}
\end{equation}
%
The integral over the cut-cell volume has been transformed in an area integral on its $nf_c=5$ faces using the divergence theorem. As these $k$ faces with areas $A_k$ and normals out of the control volume $\hat{\mathbf{n}}_{ii,k}$ are planar by construction, the method for evaluation of their mean diffusive fluxes $\left( - \rho D_\alpha \boldsymbol{\nabla} Y_\alpha \right)_k$ will define the spatial accuracy of the discretization. A centroid to centroid (i.e. $\Delta x_{cc}$ in figure~\ref{Fig:figure_2}a) finite differences and linear interpolation are used to approximate $\boldsymbol{\nabla} Y_\alpha$ and $\rho D_\alpha$ for each face belonging to the gas phase. Also, a normal probe approach~\cite{balaras_2004} is employed to sample information from the fluid to define fluxes in boundary cut-faces.
%
\begin{figure}[h]
   \centering
   \includegraphics[trim = 30mm 42mm 35mm 45mm, clip,width=1\linewidth]{./figures/SketchFig2.pdf}
   \put(-335,-5){(a)}
   \put(-95,-4){(b)}
   \caption{Sketch of cut-cell and faces: (a) Cut-cell $ii$ surrounded by gas phase regular and cut-faces (3-5), and boundary cut-faces (1-2).  (b) Interpolation sketch for wall modeled immersed boundary reconstruction of cut-face velocities.}
   \label{Fig:figure_2}
\end{figure}
%

>>>>

% Authors, for the paper (add full first names)
\Author{Marcos Vanella $^1$, Kevin McGrattan $^1$, Randall McDermott $^1$, Glenn Forney $^1$, William Mell $^2$, Emanuele Gissi $^3$, and Paolo Fiorucci $^4$}

% Authors, for metadata in PDF
\AuthorNames{Marcos Vanella, Kevin McGrattan, Randall McDermott, Glenn Forney, William Mell, Emanuele Gissi, and Paolo Fiorucci}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad National Institute of Standards and Technology, Gaithersburg, Maryland, USA \\
$^{2}$ \quad U.S. Forest Service, Seattle, Washington, USA \\
$^{3}$ \quad Corpo nazionale dei Vigili del fuoco, Savona, Italy \\
$^{4}$ \quad CIMA Research Foundation, Savona, Italy}

@@

\abstract{A method for the large-eddy simulation (LES) of wildfire spread over complex terrain is presented. In this scheme, a cut-cell immersed boundary method (CC-IBM) is used to render the complex terrain, defined by a tessellation, on a rectilinear Cartesian grid. Discretization of scalar transport equations for chemical species is done via a finite volume scheme on cut-cells defined by the intersection of the terrain geometry and the Cartesian cells. Momentum transport and heat transfer close to the immersed terrain are handled using dynamic wall models and a direct forcing immersed boundary method. Further, atmospheric wind conditions are specified using a mean forcing concept. Additionally, three basic approaches have been explored to model fire spread: (1) representing the vegetation as a collection of Lagrangian particles, (2) representing the vegetation as a semi-porous boundary, and (3) representing the fire spread using a level set method in which the fire spreads as a function of terrain slope, vegetation type, and wind speed. Several test and validation cases are reported to demonstrate the capabilities of this novel wildfire simulation methodology.}

@@

In the past few decades, wildland and wildland-urban interface (WUI) fires have received substantial attention due to their destructive nature and extensive cost in lives and property~\cite{thomas_2017,mcdermott_2019,richards_2020}. A large body of literature can be found on the subject, as seen in various review articles~\cite{Papadopoulos_2011,Bakhshaii_2019,mcdermott_2019}. Due to the complex nature of fire, limited computing resources, and the needs of planners and first responders, most models of wildfires have historically relied on simplified field-tested rules and correlations. Among these, the rate of spread model of Rothermel~\cite{Rothermel:1972} assumes quasi-steady state surface fire conditions and takes as inputs fuel properties, wind, terrain slope, and moisture that can be measured {\em in situ}. Fuel models have been cataloged~\cite{Anderson:1982}, and their parameters defined so that model users can select and combine the most appropriate listed fuels. Rothermel's rate of spread formula is still widely used, and has been implemented in several simulation tools~\cite{Finney:FARSITE,Bova:IJWF2015,mcgratta_2013,Coen:2,Coen:2015,LAUTENBERGER_2013,Coen:2013,Mandel:2009,Mandel:2011,Mandel:2014,Kochanski:2016}, enabling a range of planning and operational forecasting capabilities.

Wildland fire modeling typically focuses on tracking the fire front, which can have a complex shape that moves and deforms based on local conditions like fuel loading, moisture content, wind, and terrain. The fire front requires a mathematical representation. The Lagrangian approach is based on a polygonal mesh of control markers that represents the front~\cite{Finney:FARSITE,Bova:IJWF2015}. On the other hand, an Eulerian representation allows for a fixed two-dimensional mesh on which the fire front is defined implicitly by a scalar field. This scalar field is called the level set function and numerical methods that solve for it in time are called Level Set Methods (LSM)~\cite{Sethian:1999,Osher:2006}. Some wildfire solvers that use LSM to track the fire front are described in references~\cite{coen_2013,Bova:IJWF2015,mcgratta_2013,LAUTENBERGER_2013}. A practical comparison of the two techniques as implemented in the Fire Dynamics Simulator (FDS)~\cite{Mell:IJWF2007} and FARSITE~\cite{Finney:FARSITE} is provided in reference~\cite{Bova:IJWF2015}.

Because the spread rate of the fire front depends in part on the local wind conditions, a well resolved wind field over complex terrain should improve the accuracy of the overall model. The de facto physics-based technique for practical simulation of wind is Large Eddy Simulation (LES)~\cite{coen_2013,Mell:IJWF2007}. For fire simulation over complex terrain, a range of techniques are being developed, in particular, in the coupling of the fire and the wind. For fast simulation of outdoor fires, an immersed boundary method has been interfaced with a Level Set scheme~\cite{Bova:IJWF2015} and the Rothermel fire spread model. The wind speed input for Rothermel's model results from the local velocity over the terrain which is affected by fire buoyancy and flow evolution. These are in turn also influenced by wind atmospheric conditions, accounted for within simulations using a mean wind forcing concept. This is a first step at implementing a level set approach in a LES wind calculation with combustion. It is recognized that, in the empirically-based development of the Rothermel formula, the wind speed was not influenced by the local buoyancy induced flow.

In the following section the mathematical model for fire spread over complex terrain is described. The description of the terrain discretization is provided in section~\ref{sec:terraindisc}. The various fire spread methods are described in section~\ref{sec:firespread}, and the atmospheric boundary conditions in section~\ref{sec:wind}. Some flat terrain simulations are compared to experimental data, and a complex terrain simulation is compared to an actual wildfire in section~\ref{sec:numexp}.

\section{Mathematical Model} \label{sec:matmodel}

@@

Consider a mixture composed of $N$ chemical species $\alpha$, moving on a fixed point $\mathbf{x}$ in space with a mass weighted average velocity $\mathbf{u}(\mathbf{x},t)$. If $\rho(\mathbf{x},t)$ is the mixture density, the mass fraction for species $\alpha$ is $Y_\alpha = \rho_\alpha / \rho$, where $\rho_\alpha(\mathbf{x},t)$ is the species mass density and $\rho = \sum \rho_\alpha$. The scalar transport and momentum equations take the form:
\begin{eqnarray}
   \frac{\partial \rho Y_\alpha}{ \partial t} + \nabla \cdot ( \rho Y_\alpha  \mathbf{u} ) &=& - \nabla \cdot \mathbf{J_{d}}_\alpha + \dot{m}_\alpha'''  +
    \dot{m}_{b,\alpha}'''\; , \; \alpha=1,\dots,N \label{eqn:spectran} \\
    \frac{\partial \mathbf{u}}{\partial t} - \mathbf{u} \times \boldsymbol{\omega} + \nabla H - \tilde{p} \, \nabla \left( 1/\rho\right) &=&
    \frac{1}{\rho} \left[ (\rho-\rho_0) \mathbf{g} + \mathbf{f}_{b} + \nabla \cdot \boldsymbol{\tau} \right] \label{eqn:momtran}
\end{eqnarray}
where $\mathbf{J_{d}}_\alpha=- \rho D_\alpha \boldsymbol{\nabla} Y_\alpha$ are the diffusive fluxes for component $\alpha$. Fick's Law for binary diffusion respect to a background species is assumed, $D_\alpha$ is the diffusivity of $\alpha$ respect to the background species.  To enforce realizability, $\sum_\alpha \mathbf{J_{d}}_\alpha = 0$, we lump errors in diffusive transport into the most abundant species locally \cite{McDermott:2015}. The volumetric combustion source term for species $\alpha$ is $\dot{m}_\alpha'''(\mathbf{x},t)$  and $\dot{m}_{b,\alpha}'''$ is the contribution to species $\alpha$ from subgrid particle gasification. In the momentum equation the head is $H=|\mathbf{u}|^2/2 + \tilde{p}/\rho$, $\tilde{p}$ is the perturbation pressure, and $\rho_0(z)$ is an equlibrium density.  The gravity vector is $\mathbf{g}=(0,0,g_z)$, $\mathbf{f}_{b}$ is the particle force term, and $\boldsymbol{\tau}$ is a tensor accounting for molecular and subgrid turbulent stresses parameterized by an effective eddy viscosity model. This last model is employed by default in large-eddy simulation (LES) of outdoor fires. Details of the model equations and their derivation are found in \cite{mcgratta_2013tr}.

Conservation of energy is achieved by forcing flow field to obey the following thermodynamic divergence constraint:
\begin{eqnarray}
    ( \nabla \cdot \mathbf{u} )^{th} &=&
    \left[ \frac{1}{\rho c_p T} - \frac{1}{\bar{p}} \right]
    \frac{\partial \bar{p}}{\partial t} + \frac{w \rho_0 g_z}{\rho c_p T} \nonumber \\
    &+& \frac{1}{\rho c_p T} \left[ \dot{q}''' + \dot{q}_b''' - \nabla \cdot \dot{\mathbf{q}}'' - \mathbf{u} \cdot \nabla (\rho h_s) \right] \nonumber \\
    &+& \frac{1}{\rho} \sum_\alpha \left( \frac{\overline{W}}{W_\alpha} - \frac{h_{s,\alpha}}{c_p T} \right) \left[ \dot{m}_\alpha'''  + \dot{m}_{b,\alpha}''' - \nabla \cdot (\mathbf{J_{d}}_\alpha) - \mathbf{u} \cdot \nabla (\rho Y_\alpha) \right] \label{eq:divth}
\end{eqnarray}
where $h_s$ is the local sensible enthalpy, $\dot{q}''', \dot{q}_b'''$ are heat release rates due to combustion and particle gasification, $\dot{\mathbf{q}}''$ is the heat flux, sum of conduction, convection and radiation, and $T$ is the local temperature. The mixture specific heat at constant pressure and molecular weight are $c_p=\sum_{\alpha =1}^N{c_{p,\alpha} Y_\alpha}$ and $\overline{W}=\left(\sum_{\alpha =1}^N{Y_\alpha /W_\alpha} \right)^{-1}$ respectively, built from the individual species counterparts as defined. Finally, $w,\rho_0,g_z$ on the stratification term are respectively, the local vertical velocity, density for standard conditions (depends on vertical coordinate), and vertical gravity acceleration. This thermodynamic divergence is derived from and acts as a proxy for the sensible enthalpy evolution equation~\cite{mcdermo_2014}. Equation (\ref{eq:divth}) is derived by factoring the divergence from the sensible enthalpy equation and applying the ideal gas law.  Given a background pressure and a local mass density the local temperature is computed from the ideal gas law as $T = \bar{p} \overline{W} / (\rho R) $.

@@

In figure~\ref{Fig:figure_1}a, a terrain is represented by its surface triangulation within an FDS cartesian mesh. Some cartesian cells and faces belonging to this mesh are transversed by the terrain surface. The remainder polyhedra and polygons that lay on the gas side of this intersected cartesian entities are called cut-cells and cut-faces. They discretize the fluid domain in the region surrounding the solid terrain. A detail of this unstructured mesh is shown in figure~\ref{Fig:figure_1}b. The resulting cut-cells and cut-faces are of varied sizes and shapes, and situations where more than one cut-cell per cartesian cell are not uncommon. As described later, small cells pose a temporal stability constraint in explicit time integration schemes that needs to be addressed.
It is noted that a two level of refinement grid hierarchy emerges. The coarse level is defined by the Cartesian entities, whereas the fine level is defined by the cut-cell or unstructured counterparts. Methods that solve discrete model equations on these grids are called cut-cell or embedded boundary methods~\cite{berger_2016}. Reliably defining the cut-cell mesh geometric properties and topology is a complex problem on itself, omitted here in order to keep the discussion succinct.
\begin{figure}[ht]
   \centering
   \includegraphics[trim = 12mm 42mm 8mm 25mm, clip,width=1\linewidth]{./figures/SketchFig1.pdf}
   \put(-305,-5){(a)}
   \put(-95,-4){(b)}
   \caption{Sketch of regular Cartesian and cut-cell grids around terrain: (a) An unstructured cut-cell region is defined on the gas phase side around a terrain $\mathbf{T}$. (b) Detail showing a slice representative of cut-cells and regular gas cells, terrain triangulation and two refinement level interpretation.}
   \label{Fig:figure_1}
\end{figure}

@@

The equations for chemical species transport are discretized using the finite volume method (FV)~\cite{eymard_2000,leveque_2002}. See figure~\ref{Fig:figure_2}a. Integrating equation~\eqref{eqn:spectran} over a cut-cell $ii$ with control volume $\Omega_{ii}$:
\begin{equation}
 \int_{\Omega_{ii}} {\frac{\partial \rho Y_\alpha}{\partial t}} d \Omega + \int_{\Omega_{ii}} { \boldsymbol{\nabla} \cdot  \left(  \rho Y_\alpha \mathbf{u} \right)
      } d \Omega  = -\int_{\Omega_{ii}} { \boldsymbol{\nabla} \cdot \left(  \mathbf{J_{d}}_\alpha  \right)  } d \Omega + \int_{\Omega_{ii}} {( \dot{m}_\alpha'''+\dot{m}_{b,\alpha}''') } d \Omega \label{eq:intconvdiff}
\end{equation}
For a control volume that does not change with time, the time derivative and source terms are approximated by
\begin{eqnarray}
  \int_{\Omega_{ii}} {\frac{\partial \rho Y_\alpha}{\partial t}} d \Omega & = & \frac{\partial}{\partial t} \int_{\Omega_{ii}} {\rho Y_\alpha} d \Omega
  = \frac{\partial \widetilde{\rho \: Y_\alpha }_{ii}}{\partial t} V_{ii} \\
  \int_{\Omega_{ii}} {( \dot{m}_\alpha''' +\dot{m}_{b,\alpha}''')} d \Omega & = & (\widetilde{ \dot{m}_\alpha''' }_{ii}+ \widetilde{\dot{m}_{b,\alpha}'''}_{ii}) V_{ii} \label{eq:intcons}
\end{eqnarray}
where $V_{ii}$ is the volume of cell $ii$ and tildes imply cell averages. In the following the notation is simplified by dropping tildes, observing that quantities are cell or face averaged. Consider the FV discretization of the diffusive term (eqn.~\eqref{eq:intconvdiff}) on cut-cell $ii$ of figure~\ref{Fig:figure_2}a:
\begin{equation}
\int_{\Omega_{ii}} { \boldsymbol{\nabla} \cdot \left(  \mathbf{J_{d}}_\alpha  \right)  } d \Omega =
    \int_{\partial \Omega_{ii}} { \left( - \rho D_\alpha \boldsymbol{\nabla} Y_\alpha \right) \cdot \hat{\mathbf{n}}_{ii} } \: d \partial \Omega = \sum^{nf_c}_{k=1}
    \left( - \rho D_\alpha \boldsymbol{\nabla} Y_\alpha \right)_k \cdot \hat{\mathbf{n}}_{ii,k} \: A_k \label{eq:discfvdiffcc}
\end{equation}
The integral over the cut-cell volume has been transformed in an area integral on its $nf_c=5$ faces using the divergence theorem. As these $k$ faces with areas $A_k$ and normals out of the control volume $\hat{\mathbf{n}}_{ii,k}$ are planar by construction, the method for evaluation of their mean diffusive fluxes $\left( - \rho D_\alpha \boldsymbol{\nabla} Y_\alpha \right)_k$ will define the spatial accuracy of the discretization. A centroid to centroid (i.e. $\Delta x_{cc}$ in figure~\ref{Fig:figure_2}a) finite differences and linear interpolation are used to approximate $\boldsymbol{\nabla} Y_\alpha$ and $\rho D_\alpha$ for each face belonging to the gas phase. Also, a normal probe approach~\cite{balaras_2004} is employed to sample information from the fluid to define fluxes in boundary cut-faces.
\begin{figure}[ht]
   \centering
   \includegraphics[trim = 30mm 42mm 35mm 45mm, clip,width=1\linewidth]{./figures/SketchFig2.pdf}
   \put(-335,-5){(a)}
   \put(-95,-4){(b)}
   \caption{Sketch of cut-cell and faces: (a) Cut-cell $ii$ surrounded by gas phase regular and cut-faces (3-5), and boundary cut-faces (1-2).  (b) Interpolation sketch for wall modeled immersed boundary reconstruction of cut-face velocities.}
   \label{Fig:figure_2}
\end{figure}
Similarly, the discretization of the advective term is:
\begin{equation}
    \int_{\Omega_{ii}} { \boldsymbol{\nabla} \cdot  \left(  \rho Y_\alpha \mathbf{u} \right)} d \Omega =
     \int_{\partial \Omega_{ii}} { \left( \rho Y_\alpha \mathbf{u} \right) \cdot \hat{\mathbf{n}}_{ii} } \: d \partial \Omega =
     \sum^{nf_c}_{k=1} \left( \rho Y_\alpha \mathbf{u} \right)_k \cdot \hat{\mathbf{n}}_{ii,k} \: A_k \label{eq:discfvadvcc}
\end{equation}
